A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 MS.tex
Cross references between files now should be working

\usepackage{xr}
\externaldocument{figures}

@@

We reasoned that if dispersal out of the recapture area was reducing the
LBL below the true lifespan then there should be a positive correlation
between the spatial extent of the recapture zone and LBL. We found no
such pattern (Fig. S\#), although there was a positive correlation
between LBL and trap density (Fig. S\#).

@@

The MRR experiments included information on whether mosquitoes were
pre-fed with sugar (41 time series), blood (71), both (4) or
alternatively unfed (116). We estimate that female mosquitoes that were
fed on sugar pre-release lived on average for 0.7 days longer than those
that were not fed (Fig. S\#; a pattern that was consistent across the
genera). There were insufficient males that were either fed or unfed
with sugar prior to release to make a meaningful comparison. Females
that were blood-fed prior to release on average lived 1.5 days longer
than those who were not fed for \emph{Aedes} but this trend was reversed
for \emph{Anopheles} meaning that there was little difference overall.

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
(\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}) individually (Fig. S\#).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the Methods (Fig. S\#). For ten species, we had enough data from both
species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - (Table S\#) there was a significant difference in the
time-series level LBLs.

@@

In Fig. \ref{fig:mrr_elpd}, we compare the performance of the six models for describing
lifespan in MRR studies of 33 species using K-fold cross-validation. We
categorised the evidence for age-dependent mortality in each species
according to the performance of the five age-dependent models versus the
exponential: `+' indicated that all age-dependent models outperformed
the exponential; `?' indicated that the exponential outperformed one or
more age-dependent models; and `-' indicated that the exponential
performed at least as well as all other models. Overall, we estimated
that there were 8 `+' species, where age-dependent mortality fit the
data better; 11 `?' species where the evidence was mixed; and 14 species
where constant mortality models performed at least as well. The species
where age-dependent mortality best fit the data included the major
vector of dengue fever, Zika and chikungunya, \emph{Ae. Aegypti}. These
studies also tended to include multiple release MRR studies which, on
average, were conducted over a longer period of time than the others,
which may be why we failed to detect age-dependence in the latter (Fig
S\#).

@@

MRR experiments are known to produce downwardly-biased estimates of
lifespan. Lab experiments have demonstrated that marking can negatively
impact survival (Verhurst et al., 2013; Dickens and Brant, 2014),
resulting in artificially depressed survival. MRR studies typically
cannot differentiate between a mosquito dying and dispersal from the
study area meaning that lifespan will be underestimated. In this study,
we found a positive correlation between lifespan estimates and the
density of traps, indicating that better trapping coverage likely raises
estimates towards their real value. We conducted an \emph{in silico}
Monte Carlo study to determine how accurately we could estimate mosquito
lifespan given study parameters in an ideal MRR experiment, where the
assumptions of no emigration and harmless marking are fully satisfied
(see SOM for full details). This work indicated that for many of the
experiments, the short study lengths or typical numbers of mosquitoes
released, results in considerable uncertainty in lifespan estimates (Fig
S\#). This indicates that statistical power can be substantially
increased by pooling data across experiments as we did using a Bayesian
hierarchical model.

@@

longer gonotrophic cycle (Fig. S\#). The distributions we used to

@@

By applying a common method to analysing all studies in our databases,
it is possible that we may have missed patterns of mortality that would
have been evident from using a more bespoke approach. As our \emph{in
silico} analysis of MRR experiments indicates, however, the
overdispersed data from single experiments results in high measurement
error (Fig. S\#). By applying different methods to each study, this
could lead us to falsely detect patterns when none are present, and we
prefer a pooled approach.

@@

We conducted a power analysis of MRR experiments (Fig. S\#) to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
S\#) for case study populations with three different levels of
senescence (Fig. S\#). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. S\#) but is
insensitive to release size (Fig. S\#). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study length for
experiments included in our analysis was 10 days (Table S\#) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations (Styer et al., 2007; Dawes et al., 2009).
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. S\#). Our analysis
indicates that females outlive male mosquitoes by approximately 0.9 days
(Fig. \#), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquitoes.

@@

There is considerable evidence from the laboratory that temperature
modulates mosquito ecology and behaviour (Matt Thomas references). The
locations and times of year over which the MRR studies were conducted
encompassed a large range of average air temperatures, from
approximately 10 \textsuperscript{o}C to 35 \textsuperscript{o}C and,
within this, we determined no relationship between lifespan and
temperature across all time series (Fig. 4) or, for any of the species
with the most data (Fig. S\#). It is possible that by considering a raw
average of air temperature across the month, this ignored other
temperature-dependent factors of importance for longevity, such as
temperature range. It is also possible that by ignoring the effects of
rainfall (the historical data on rainfall is less likely to be reliable
for a given location), that this masked a more complex interaction
between longevity and temperature. The observed laboratory relationship
between lifespan and temperature, however, may not be as robust in the
field if mosquitoes adjust their behaviours (such as, by seeking shade)
in reaction to changes in temperature. More work exploring the
relationship between mosquito ecology and temperature in semi-field
experiments may be useful in probing these interactions further.

The importance of vector mortality for disease transmission has long been recognised,
since even before 1957, when George Macdonald formulated the
now famous Ross-Macdonald equation of R\textsubscript{0} for malaria. Indeed, the recent declines in malaria prevalence in Sub-Saharan Africa were likely due to upscaling of interventions (insecticide-treated bednets and indoor residual spraying) that aim to reduce mosquito lifespan (Bhatt et al., 2015). Worryingly, resistance to pyrethroids, the only class of insecticide used in current insecticide-treated bednets and likely the only product to come to market in the near
future, has been determined to be widespread and increasing in intensity
across Sub-Saharan Africa (WHO reference 2018). This alarming trend
highlights the need for continued MRR and dissection-based studies to
monitor the effectiveness of bednets and determine whether more
expensive alternatives, such as nets incorporating piperonyl butoxide be
deployed. It also emphasises the need for investment in new tools for
real time monitoring of mosquito populations. In this work, we have used modern statistical methods to synthesise precious field data conducted by entomologists past and present, to produce lower bound estimates of mosquito lifespan. In recent years, there has been considerable funding allocated to molecular and genomic research
into mosquitoes that strengthens existing interventions and suggest
novel control strategies. Without commensurate funding allocated to
applied vector ecology, our lack of knowledge in this area threatens our
opportunity to capitalise on molecular advances and potentially hinders
our ability to control of mosquito-borne disease.

@@

The authors would like to thank the following for useful conversations
throughout the course of this work: Austin Burt, Thomas Churcher, Steve
Lindsay, Mike Bonsall and Ellie Sherrard-Smith.

>>>>

\usepackage{xr}
\externaldocument{figures}
\externaldocument[S-]{supplementary_materials}

@@

We reasoned that if dispersal out of the recapture area was reducing the
LBL below the true lifespan then there should be a positive correlation
between the spatial extent of the recapture zone and LBL. We found no
such pattern (Fig. \ref{S-fig:mrr_lifeSpanVsRange}), although there was a positive correlation
between LBL and trap density (Fig. \ref{S-fig:mrr_lifeSpanVsTrapDensity}).

@@

The MRR experiments included information on whether mosquitoes were
pre-fed with sugar (41 time series), blood (71), both (4) or
alternatively unfed (116). We estimate that female mosquitoes that were
fed on sugar pre-release lived on average for 0.7 days longer than those
that were not fed (Fig. \ref{S-fig:mrr_female_blood_sugar}; a pattern that was consistent across the
genera). There were insufficient males that were either fed or unfed
with sugar prior to release to make a meaningful comparison. Females
that were blood-fed prior to release on average lived 1.5 days longer
than those who were not fed for \emph{Aedes} but this trend was reversed
for \emph{Anopheles} meaning that there was little difference overall.

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
(\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}) individually (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}).

@@

To compare the two methods, we converted numbers of gonotrophic cycle
(physiological age) into lifespan (chronological age) as described in
the SOM using estimates of gonotrophic cycle duration (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw}). For ten species, we had enough data from both
species to make a comparison, and there was a positive correlation (not
statistically significant; Pearson correlation $\rho=0.42$, $n=10$, $p=0.23$)
between the two measures (Fig. \ref{fig:comparison}), and in only one case -- for \emph{A.
darlingi} - (Table S\#) there was a significant difference in the
time-series level LBLs.

@@

In Fig. \ref{fig:mrr_elpd}, we compare the performance of the six models for describing
lifespan in MRR studies of 33 species using K-fold cross-validation. We
categorised the evidence for age-dependent mortality in each species
according to the performance of the five age-dependent models versus the
exponential: `+' indicated that all age-dependent models outperformed
the exponential; `?' indicated that the exponential outperformed one or
more age-dependent models; and `-' indicated that the exponential
performed at least as well as all other models. Overall, we estimated
that there were 8 `+' species, where age-dependent mortality fit the
data better; 11 `?' species where the evidence was mixed; and 14 species
where constant mortality models performed at least as well. The species
where age-dependent mortality best fit the data included the major
vector of dengue fever, Zika and chikungunya, \emph{Ae. Aegypti}. These
studies also tended to include multiple release MRR studies which, on
average, were conducted over a longer period of time than the others,
which may be why we failed to detect age-dependence in the latter (Fig
\ref{S-fig:mrr_mcPowerAnalysis_senescence}).

@@

MRR experiments are known to produce downwardly-biased estimates of
lifespan. Lab experiments have demonstrated that marking can negatively
impact survival (Verhurst et al., 2013; Dickens and Brant, 2014),
resulting in artificially depressed survival. MRR studies typically
cannot differentiate between a mosquito dying and dispersal from the
study area meaning that lifespan will be underestimated. In this study,
we found a positive correlation between lifespan estimates and the
density of traps, indicating that better trapping coverage likely raises
estimates towards their real value. We conducted an \emph{in silico}
Monte Carlo study to determine how accurately we could estimate mosquito
lifespan given study parameters in an ideal MRR experiment, where the
assumptions of no emigration and harmless marking are fully satisfied
(see SOM for full details). This work indicated that for many of the
experiments, the short study lengths or typical numbers of mosquitoes
released, results in considerable uncertainty in lifespan estimates (Fig.
\ref{S-fig:mrr_mcPowerAnalysis}). This indicates that statistical power can be substantially
increased by pooling data across experiments as we did using a Bayesian
hierarchical model.

@@

longer gonotrophic cycle (Fig. \ref{S-fig:dissection_gonotrophicCycleRaw_MRRVsLab}). The distributions we used to

@@

By applying a common method to analysing all studies in our databases,
it is possible that we may have missed patterns of mortality that would
have been evident from using a more bespoke approach. As our \emph{in
silico} analysis of MRR experiments indicates, however, the
overdispersed data from single experiments results in high measurement
error (Fig. \ref{S-fig:mrr_mcPowerAnalysis}). By applying different methods to each study, this
could lead us to falsely detect patterns when none are present, and we
prefer a pooled approach.

@@

We conducted a power analysis of MRR experiments to determine
whether typical experimental characteristics could detect senescence.
Here we calculated the power of a maximum likelihood estimator of the
`senescence parameter' $\beta$ of the Gompertz survival function (see Table
S\#) for case study populations with three different levels of
senescence (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}A). This analysis indicated that power to detect
senescence strongly depends on study length (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}B) but is
insensitive to release size (Fig. \ref{S-fig:mrr_mcPowerAnalysis_senescence}C). Clements and Patterson (1981)
conducted a meta-analysis of MRR and dissection-based field experiments
and found evidence of an increasing risk of mortality hazard with age
that is similar in magnitude to that of the `mild' case considered
above. For this case, detecting senescence with a power of 80\% requires
a study length of at least 18 days. Since the median study length for
experiments included in our analysis was 10 days (Table S\#) this could
partly explain our failure to detect senescence at the species level. A
number of experiments have found evidence of age-dependence in
laboratory populations (Styer et al., 2007; Dawes et al., 2009).
However, the artificially benign environment of the laboratory means
mosquitoes live considerably longer than in the wild, where they may die
because of exogenous factors, before the effects of physiological
decline have had time to manifest. Field experiments have also found
evidence for age-dependent mortality. Harrington et al. (2008) conducted
a field experiment where mosquitoes reared under laboratory conditions
were marked and released at different ages. Analysis of the resultant
MRR time-series indicated that mosquito mortality increases with age at
release. It is possible, however, that this field experiment suffers
from the same biases as laboratory-based approaches, because the
released mosquitoes were often of ages considerably higher (up to 20
days) than typical estimates of wild mosquito lifespan.

As ethical concerns of contributing to disease burden are more often
considered, it is now less common for MRR experiments to release female
mosquitoes versus males than historically (Fig. \ref{S-fig:mrr_sexReleasesOverTime}). Our analysis
indicates that females outlive male mosquitoes by approximately 0.9 days
(Fig. \ref{fig:mrr_sexDifferences_without_sugar_nor_blood}), meaning that differences between the sexes may exist for
other ecological parameters determinable by MRR. This suggests that
continued field entomological work on contained releases of mosquitoes
in semi-field sites or large microcosms may be a valuable source of
information on female mosquitoes.

@@

There is considerable evidence from the laboratory that temperature
modulates mosquito ecology and behaviour (Matt Thomas references). The
locations and times of year over which the MRR studies were conducted
encompassed a large range of average air temperatures, from
approximately 10 \textsuperscript{o}C to 35 \textsuperscript{o}C and,
within this, we determined no relationship between lifespan and
temperature across all time series (Fig. \ref{fig:mrr_temperature}) or, for any of the species
with the most data (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}). It is possible that by considering a raw
average of air temperature across the month, this ignored, more complex, interactions between temperature and lifespan. It is also possible that by ignoring the effects of
rainfall (the historical data on rainfall is less likely to be reliable
for a given location), that this masked a more complex interaction
between longevity and temperature. The observed laboratory relationship
between lifespan and temperature, however, may not be as robust in the
field if mosquitoes adjust their behaviours (such as, by seeking shade)
in reaction to changes in temperature. More work exploring the
relationship between mosquito ecology and temperature in semi-field
experiments may be useful in probing these interactions further.

In this work, we have used modern statistical methods to synthesise precious field data conducted by entomologists past and present, to produce lower bound estimates of mosquito lifespan. The importance of vector mortality for disease transmission has long been recognised, however,
since even before 1957, when George Macdonald formulated the
now famous Ross-Macdonald equation of R\textsubscript{0} for malaria. Indeed, the recent declines in malaria prevalence in Sub-Saharan Africa were likely due to upscaling of interventions (insecticide-treated bednets and indoor residual spraying) that aim to reduce mosquito lifespan (Bhatt et al., 2015). Worryingly, resistance to pyrethroids, the only class of insecticide used in current insecticide-treated bednets and likely the only product to come to market in the near
future, has been determined to be widespread and increasing in intensity
across Sub-Saharan Africa (WHO reference 2018). This alarming trend
highlights the need for continued MRR and dissection-based studies to
monitor the effectiveness of bednets and determine whether more
expensive alternatives, such as nets incorporating piperonyl butoxide be
deployed. It also emphasises the need for investment in new tools for
real time monitoring of mosquito populations. In recent years, there has been considerable funding allocated to molecular and genomic research
into mosquitoes that strengthens existing interventions and suggest
novel control strategies. Without commensurate funding allocated to
applied vector ecology, our lack of knowledge in this area threatens our
opportunity to capitalise on molecular advances and potentially hinders
our ability to control of mosquito-borne disease.

@@

The authors would like to thank the following for useful conversations
throughout the course of this work: Austin Burt, Mike Bonsall, Thomas Churcher, Steve
Lindsay and Ellie Sherrard-Smith.

### 2018-11-23 supplementary_materials.tex
Changed gir attributes
Cross references between files now should be working

\documentclass[12pt]{article}
  \usepackage{geometry}
 \usepackage[round]{natbib}
 \usepackage{graphicx}
 \geometry{a4paper}
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{authblk}
  \usepackage[running]{lineno}
  \usepackage{setspace}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{chngpage}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage[hidelinks]{hyperref}
 \usepackage[textsize=tiny]{todonotes}
\usepackage{pdfpages, caption}
\usepackage{listings}
\usepackage{minted}
\usepackage{framed}
\usepackage{mdframed}

@@

@@

@@

>>>>

\documentclass[12pt]{article}
  \usepackage{geometry}
 \usepackage[round]{natbib}
 \usepackage{graphicx}
 \geometry{a4paper}
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{authblk}
  \usepackage[running]{lineno}
  \usepackage{setspace}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{chngpage}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage[hidelinks]{hyperref}
 \usepackage[textsize=tiny]{todonotes}
\usepackage{pdfpages, caption}
\usepackage{listings}
\usepackage{minted}
\usepackage{framed}
\usepackage{mdframed}
\usepackage[section]{placeins}

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=1\textwidth]{./Figure_files/mrr_sexReleasesOverTime.pdf}}
	\caption{\textbf{The number of MRR time-series by date and sex of released mosquitoes.} The numbers represent by-decade totals of the 232 time-series we used in the main analysis.}
	\label{fig:mrr_sexReleasesOverTime}
\end{figure}

@@

\begin{figure}[ht]
	\centerline{\includegraphics[width=1.25\textwidth]{./Figure_files/mrr_mcPowerAnalysis.pdf}}
	\caption{\textbf{The average error in predicting mean lifespan as a function of study length (left) and number of released mosquitoes (right) in a Monte Carlo analysis.} For each parameter set we generated 400 simulated data series using the negative binomial sampling model with an exponential survival function as described in the text, and estimated the mean lifespan using maximum likelihood. The error bars show the standard deviation in the prediction error.}
	\label{fig:mrr_mcPowerAnalysis}
\end{figure}

\begin{figure}[ht]
	\centerline{\includegraphics[width=1.25\textwidth]{./Figure_files/mrr_mcPowerAnalysis_senescence.pdf}}
	\caption{\textbf{The statistical power to detect senescence for A. three case study populations as a function of B. study length and C. number released.} For each parameter set we generated 500 simulated data series using the negative binomial sampling model with a gompertz hazard function ($\lambda(t) = \alpha e^{\beta t}$), and estimated the $\beta$ parameter using maximum likelihood and tested the null hypothesis $\beta=0$ (constant mortality risk) against the alternative $\beta>0$ using an likelihood ratio test with a 5\% test size. The error bars show the standard deviation in the power for each parameter set. The recapture probability parameters ($\psi$ and $\kappa$) used were the averages of those estimated for the actual data.}
	\label{fig:mrr_mcPowerAnalysis_senescence}
\end{figure}

@@

\subsection{The statistical power of MRR studies}\label{sec:mrr_discussion_MonteCarlo}
Consider an ideal MRR experiment, where the assumptions of no emigration and harmless marking are fully satisfied. With these data, how accurately can we estimate mosquito lifespan? To address this question we generated artificial MRR data and attempted to estimate the (known) parameters by maximum likelihood (a `Monte Carlo' analysis). Specifically, we simulated releases of $N$ mosquitoes that are monitored for $m$ days in each case, and determined how statistical power to estimate lifespan depends on these two parameters (Figure \ref{fig:mrr_mcPowerAnalysis}). In order to focus on estimating lifespan, we assumed the mosquitoes experience constant mortality, and recaptures follow the negative binomial sampling model (eqn. \ref{eq:NB}) where the recapture probability parameters ($\psi$ and $\kappa$) were the averages of those estimated for the actual data.

Unsurprisingly, the error in predicting lifespan declines as the duration of study is increased (Figure \ref{fig:mrr_mcPowerAnalysis}A). However once a study length is much longer than the lifespan of the majority of mosquitoes, the predictive power cannot be improved by extending the study duration. The effect of increasing release size (while holding study length constant) is similarly a case of diminishing returns (Figure \ref{fig:mrr_mcPowerAnalysis}B). For the parameters we use, there are significant gains in accuracy from releasing $1,000$ rather than $100$ mosquitoes, but modest gains from releasing $10,000$ rather than $1000$ mosquitoes.

We also conducted a power analysis for the detection of senescence in MRR experiments (Figure \ref{fig:mrr_mcPowerAnalysis_senescence}). Here we calculated the power of a maximum likelihood estimator of the `senescence parameter' $\beta$ of the Gompertz survival function (see Table \ref{tab:mrr_survivalDescription}) for case study populations with three different levels of senescence (Figure \ref{fig:mrr_mcPowerAnalysis_senescence}A.). These indicate that the power to detect senescence is strongly dependent on study length (Figure \ref{fig:mrr_mcPowerAnalysis_senescence}B.), but is insensitive to release size (Figure \ref{fig:mrr_mcPowerAnalysis_senescence}C.). \cite{clements1981analysis} conducted a meta-analysis of MRR and dissection-based experiments and found evidence of senescence that is, at least, qualitatively similar in magnitude to that of the `mild' case considered above. For this case detecting senescence with a power of 80\% requires a study length of at least 18 days. Since the median study length for experiments included in our analysis was 10 days (Table \ref{tab:mrr_aggregateData}) this could partly explain our failure to detect senescence at the species level.
